---
title: "Non-stationary responses in anchovy (_Engraulis encrasicolus_) recruitment to
  coastal upwelling in the Southern Benguela: Revisited"
output: 
  html_document:
    code_folding: hide
date: "2024-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction




```{r load_dat, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
source("https://raw.githubusercontent.com/duncanobrien/ews-assessments/main/Code/threshold_gam.R")

upwelling <- readRDS(here::here("data/upwelling_index.rds")) %>% 
  filter(area == "total")

recruitment <- readRDS(here::here("data/pelagic_recruitment.rds")) %>% 
  filter(species == "Anchovy",
         unit == "n",
         year <= 2020)

biomass <- readRDS(here::here("data/pelagic_biomass.rds")) %>% 
  filter(species == "Anchovy",
         year <= 2020,
         indicator == "biomass") %>% 
  mutate(value = case_when(indicator == "biomass" ~ value/1000000,
                           indicator != "biomass" ~ value,
                           TRUE ~ NA_integer_))

```


```{r intro_plots, fig.height = 7, fig.width = 6.5, warning=FALSE, message=FALSE}

bio1 <- ggplot(biomass, aes(x = year, y = value, group = area, color = area)) +
  geom_path() +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "",
       y = "",
       title = "Anchovy spawner biomass",
       subtitle = "spring/summer (October/November) acoustic survey (Million tonnes)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL

rec1 <- ggplot(recruitment, aes(x = year, y = value)) +
  geom_path() +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "",
       y = "",
       title = "Anchovy recruitment",
       subtitle = "autumn/winter (May/June) acoustic survey (Billions)") +
  theme_minimal() +
  NULL

## 
upwelling_month <- upwelling %>% 
  mutate(upwelling_date = lag(year, n = 6), ##  years run from July (previous year) to June (current year). eg: 1979 = 1 July 1979 to 30 June 1980
         upwelling_year = lubridate::year(upwelling_date),
         months = lubridate::month(year)) %>% 
  filter(!is.na(upwelling_date)) 

upwelling_dat <- upwelling_month %>% ## cumulative coastal upwelling from December to March
  group_by(upwelling_year) %>%
  summarize(value = sum(value, na.rm = TRUE))

upwelling_summer_dat <- upwelling_month %>% 
  filter(!is.na(upwelling_date), 
         months %in% c(12, 1, 2, 3)) %>% ## cumulative coastal upwelling from December to March
  group_by(upwelling_year) %>%
  summarize(value = sum(value, na.rm = TRUE))

## For comparison with Lamont et al 2018 Figure 6a
# up_total <- ggplot(upwelling_dat, aes(x = upwelling_year, y = value/1000)) +
#   geom_bar(stat = "identity") +
#   scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
#   labs(x = "", y = "", 
#        title = "Upwelling index",
#        subtitle = bquote("July-June cumulative upwelling"~(m^3~s^-1~100~m^-1)~x~1000)) +
#   theme_minimal() +
#   NULL

up_summer <- ggplot(upwelling_summer_dat, aes(x = upwelling_year, y = value/1000)) +
  geom_path() +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "", y = "", 
       title = "Upwelling index",
       subtitle = bquote("December-March cumulative upwelling"~(m^3~s^-1~100~m^-1)~x~1000)) +
  theme_minimal() +
  NULL

up_summer / rec1/ bio1

```





```{r rec-plots, warning=FALSE, message=FALSE}

up_rec <- recruitment %>%
  select(year, recruits = value) %>% 
  left_join(upwelling_summer_dat %>% rename(year = upwelling_year, upwelling = value),
            join_by(year)) %>% 
  mutate(upwelling = scale(upwelling)[,1],
         recruits = scale(recruits)[,1]) %>% 
  na.omit()

cor_val <- cor(y = up_rec$recruits, x = up_rec$upwelling, method = c("pearson", "kendall", "spearman")[1])

up_rec_p <- up_rec %>% 
  pivot_longer(!year, names_to = "var", values_to = "val") 


up_rec1 <- ggplot() +
  geom_path(data = up_rec_p, aes(x = year, y = val, color = var)) +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  scale_color_manual(values = c("black", "red")) + 
  labs(x = "", y = "", 
       color = "", 
       title = "Upwelling index and Anchovy recruitment",
       subtitle = paste0("z-scored (scaled and centered), Pearson's r=", round(cor_val, 3))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
up_rec1

```


```{r lag-corr}

calculate_lags <- function(df, var, lags){
  map_lag <- lags %>% purrr::map(~purrr::partial(lag, n = .x))
  return(df %>% mutate(across(.cols = {{var}}, .fns = map_lag, .names = "{.col}_lag{lags}")))
}


corr_lags <- upwelling_month %>% 
  select(year = upwelling_year, months, upwelling = value) %>% 
  left_join(recruitment %>% select(year, recruits = value), by = join_by(year)) %>% 
  group_by(months) %>% 
  calculate_lags(var = upwelling, lags = 1L:3L) %>% 
  summarize(`lag 0` = cor(x = recruits, y = upwelling, use = "complete.obs", method = "spearman"),
            `lag 1` = cor(x = recruits, y = upwelling_lag1, use = "complete.obs", method = "spearman"),
            `lag 2` = cor(x = recruits, y = upwelling_lag2, use = "complete.obs", method = "spearman"),
            `lag 3` = cor(x = recruits, y = upwelling_lag3, use = "complete.obs", method = "spearman")) %>% 
  pivot_longer(!months, names_to = "lags", values_to = "rho") %>% 
  mutate(months = purrr::map(months, ~month.name[.x]),
         months = factor(months, levels = c("July", "August", "September",
                                            "October", "November", "December",
                                            "January", "February", "March",
                                            "April", "May", "June")))

ggplot(corr_lags, aes(x = months, y = rho, group = lags, color = lags, fill = lags)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "",
       subtitle = "",
       x = "month",
       y = "Spearman's rho",
       color = "",
       fill = "") +
  theme_minimal() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45)) +
  NULL



# acf(recruitment %>% filter(year <= 2015) %>% pull(value), plot = FALSE, lag = 1)
# acf(upwelling_summer_dat$value, plot = FALSE, lag = 1)


```
These potential linkages are statistically evident: anchovy recruitment (measured in May–June) is closely related to total spawner biomass measured 6 mo later in October–November (Pearson’s r = 0.83, p < 0.001 from 1985−2014), and adult biomass affects recruitment in the next year (Pearson’s r = 0.64, p < 0.001).

```{r fig1}
#### trying to figure out what van der Sleen did.  
upwelling_lags <- upwelling_month %>% 
  select(year = upwelling_year, months, upwelling = value) %>% 
  left_join(recruitment %>% select(year, recruits = value), by = join_by(year)) %>% 
  group_by(months) %>% 
  calculate_lags(var = upwelling, lags = 3L) %>% 
  filter(months %in% c(12, 1, 2, 3)) %>%
  pivot_longer(cols = contains("upwelling"), names_to = "var", values_to = "val") %>% 
  group_by(year, var) %>% 
  summarize(val = sum(val, na.rm = TRUE))

upwelling_lags <- bind_rows(upwelling_lags, 
                recruitment %>% filter(species == "Anchovy") %>%  select(year, var = indicator, val = value) %>% 
                  mutate(val = val)) %>% 
  filter(year >= 1985,
         year <= 2015) %>% 
  group_by(var) %>% 
  mutate(val = scale(val))

ggplot(upwelling_lags)+
  geom_path(aes(x = year, y = val, color = var)) +
  scale_color_manual(values = c("black", "gray", "red")) +
  theme_minimal() +
  labs(x = "", y = "", 
       color = "",
       title = "Comparison of lagged upwelling and recruitment") +
  theme(legend.position = "bottom") +
  NULL

upwelling_red <- expand.grid(upwelling_year = seq(from = 1979, length.out = 36),
                             lambda = seq(from = 0.00, to = .99, by = 0.01)) %>% 
  left_join(upwelling_summer_dat) %>% 
  calculate_lags(var = value, lags = 1L:3L) %>% 
  filter(upwelling_year > 1983) %>%
  pivot_longer(cols = contains("value"), names_to = "upwelling", values_to = "value") %>% 
  mutate(upwelling = gsub("value", "upwelling", upwelling),
         upwelling = gsub("_", " ", upwelling),
         upwelling = ifelse(upwelling == "upwelling", paste0(upwelling, " lag0"), upwelling)) %>% 
  group_by(lambda, upwelling) %>%
  mutate(value = scale(value)[,1],
         red = (lambda*lag(value)) + value)

ggplot(upwelling_red, aes(x = upwelling_year, y = red, group = lambda)) +
  facet_wrap(~upwelling) +
  geom_path(aes(color = lambda))

upwelling_corr <- upwelling_red %>% 
  pivot_wider(id_cols = c(upwelling_year, upwelling), names_from = lambda, values_from = red) %>% 
  left_join(recruitment %>% select(upwelling_year = year, value)) %>% 
  filter(upwelling_year > 1984) %>%
  summarize(across(!c(upwelling_year, value), ~ cor(.x, value, method = "pearson"))) %>% 
  pivot_longer(cols = !upwelling, names_to = "lambda", values_to = "value")

ggplot(upwelling_corr, aes(x = lambda, y = value, group = upwelling, color = upwelling)) +
  geom_point() +
  theme_minimal() +
  NULL


```

```{r upwelling_red}



upwelling_red <- expand.grid(upwelling_year = seq(from = 1979, length.out = 36),
                             lambda = seq(from = 0.00, to = .99, by = 0.01)) %>% 
  left_join(upwelling_summer_dat) %>% 
  filter(upwelling_year > 1983) %>%
  group_by(lambda) %>%
  mutate(value = scale(value)[,1],
         red = (lambda*scale(lag(value))[,1] + value))

ggplot(upwelling_red, aes(x = upwelling_year, y = red, by = lambda)) +
  geom_path(aes(color = lambda))

upwelling_corr <- upwelling_red %>% 
  pivot_wider(id_cols = upwelling_year, names_from = lambda, values_from = red) %>% 
  left_join(recruitment %>% select(upwelling_year = year, value)) %>% 
  filter(upwelling_year > 1984) %>%
  summarize(across(!c(upwelling_year, value), ~ cor(.x, value, method = "spearman"))) %>% 
  pivot_longer(cols = everything(), names_to = "lambda", values_to = "value")

ggplot(upwelling_corr, aes(x = lambda, y = value)) +
  geom_point() +
  theme_minimal() +
  NULL


```






```{r tgam-time}
## recruits ~ s(upwelling) + s(SSB)

up_rec_ssb <- biomass %>% 
  filter(species == "Anchovy") %>%
  pivot_wider(id_cols = year, names_from = area, names_glue = "biomass_{area}", values_from = value) %>% 
  left_join(recruitment %>% filter(species == "Anchovy") %>% select(year, recruits = value), join_by(year)) %>% 
  left_join(upwelling_summer_dat %>% rename(upwelling = value), by = join_by(year == upwelling_year)) %>% 
  mutate(upwelling = scale(upwelling)[,1],
         red = (.70*scale(lag(upwelling))[,1] + upwelling)) %>% 
  na.omit()
  
### environmental thresholds formula
cont_formula <-  as.formula(paste("recruits ~ s(red, bs='tp', k=6)"))
thresh_formula <- as.formula(paste("recruits ~ 1"))

### temporal thresholds formula
# cont_formula <-  as.formula(paste("density ~ s(date, bs='tp', k=6)"))
# thresh_formula <- as.formula(paste("density ~ 1"))

### gam
dens_gam <- compare_gam(data = up_rec_ssb,
                        cont_formula = formula(cont_formula, method = "REML"),
                        thresh_formula =  formula(thresh_formula, method = "REML"),
                        thresh.var = "biomass_west", expl.var = "red",
                        thresh.range = c(0.1, 0.9), by = 1, k = 6)

dens_gam$best

best_gam <- predict_best_gam(object=dens_gam) |> 
  dplyr::select(recruits,threshold,thresh.var,upwelling,thresh.var,
                transition,fit,ci)

```