---
title: "Causal model of S. Benguela"
author: "Scott Large"
output: 
  html_document:
    code_folding: hide
date: "2024-06-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=12)
```


```{r load_dat, message = FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(lubridate)

# install.packages("dsem")
library(dsem) ## 1.2.1
library(phylopath)

all_dat <- readRDS(here::here("data/all_dat.rds"))
benguela_nino <- readRDS(here::here("data/bni.rds"))

agulhas_gradient <- readRDS(file = here::here("data/agulhas_gradient.rds"))
pressure <- readRDS(file = here::here("data/pressure.rds"))
cap_anomaly <- readRDS(file = here::here("data/cap_anomaly.rds"))
tsa <- readRDS(file = here::here("data/tsa.rds"))


upwelling <- readRDS(here::here("data/upwelling_index.rds")) #%>% 
  # filter(area == "total")

catches <- readRDS(here::here("data/pelagic_catch.rds"))

catches_m <- readRDS(here::here("data/pelagic_catch_monthly.rds"))

recruitment <- readRDS(here::here("data/pelagic_recruitment.rds")) %>% 
  filter(unit == "n")


biomass <- readRDS(here::here("data/pelagic_biomass.rds")) %>%
  filter(indicator == "biomass")

# biomass <- readRDS(here::here("data/pelagic_biomass.rds")) %>% 
#   filter(indicator == "biomass") %>% 
#   mutate(value = case_when(indicator == "biomass" ~ value/1000000,
#                            indicator != "biomass" ~ value,
#                            TRUE ~ NA_integer_))

cal_tab <- readRDS(here::here("data/calendar_table.rds"))

ltl_dat <- readRDS(here::here("data/ltl_dat.rds"))

summer_sst <- readRDS(here::here("data/summer_sst.rds"))
```

# Introduction ---

* Ecosystems are comprised of many interacting processes. Causal models help to identify the magnitude and direction of direct and indirect effects between interacting variables. Disentangle multiple drivers of ecosystem change
* IEA as tool for confronting trade-offs between multiple EBFM objectives 
* Southern Benguela is an important and dynamic system
    + Environmental drivers:
        + Thermal spawning area - Richardson
            *  Larval anchovy foraging conditions seem to be optimal in 16-19°C water (plankton production and the consistency of copepod biomass). A positive linear relationship has been found between egg abundance and the area of 16–19°C water over the western Agulhas Bank
        + Cape Agulhas Pressure - Agenbag
            * the Agulhas Bank (AB; Figure 1) is ecologically and economically important since it comprises the major spawning grounds for numerous fish species which recruit to the west coast (Hutchings et al., 2009; Kirkman et al., 2016).
            * Efficient transport between the spawning grounds and the nursery area (Agenbag J.J. (1996). Increased wind velocity (high CAP index) over the anchovy transport area (Cape Point to Cape Columbine) is likely to result in improved transport.Hence it is thought that a high CAP anomaly for September–December should result in more efficient transport to the productive nursery grounds.
            * Decadal signal in atmospheric pressure (Roy et al 2007) seen in NCEP/NCAR reanalysis data (see pressure plot)
        + Upwelling
            * Increased wind that drives upwelling also increases surface water advection offshore, which may decrease survival of early life history stages of small pelagic species during their transport from spawning grounds to the west coast nursery area (Augustin 2017).
            * van der Lingen et al (2006) hypothesized that larger zooplankton are be more abundant when upwelling is stronger and smaller zooplankton more abundant when upwelling is weaker (van der Lingen et al., 2006), which could suggest that sardine diet is negatively influenced by upwelling.
        + Phytoplankton
            * wind-driven upwelling at discrete upwelling centers in the southern Benguela is strongly pulsed and shows greater seasonal differences, with a narrow band of high plankton productivity in the southern part, restricted to the narrow coastal zone by warm Agulhas water offshore forming a front.
            * Close relationship is to be expected between primary production and upwelling of nutrient-rich water over short time scales, it is generally assumed that long-term shifts in upwelling-favourable winds will lead to changes in upwelling and, consequently, to changes in phytoplankton production and biomass on similar scales (Hampton, 2012). However, patterns observed from in situ and satellite data provide no clear indications to suggest that such a straight-forward link between primary production and upwelling variability occurs in the Benguela, but there is evidence of substantial interannual and decade-scale changes in both phytoplankton production and upwelling at a number of times and locations in both the northern and southern Benguela sub-systems.
(Verheye et al 2016) 
        + Zooplankton
        + Small Pelagic dynamics
    + Fishing
        + Small pelagic directed and bycatch
* Here, based on conceptual models developed through the ODEMM/IEA process, we develop a novel causal model to explore how fishing and the environment influence ecosystem indicators

# Methods ---

## Model area
* *TODO* Create map with the relevant spots identified

## Data ---
**(see supplement below for more figures)**

* SST
    + Anchovy thermal spawning suitability area (Richardson et al., 1998) 
    + Standardized Cape Agulhas Pressure (CAP) anomaly. (Agenbag 1996) and Roy (2007)
    + tempsurf - Mean sea temperature (°C) near the surface (5 m depth) (Huggett 2023)

```{r remote_data}
ag_plot <- ggplot(agulhas_gradient, aes(x = date, y = sst_gradient))+
  geom_path() + 
  geom_point() +
  theme_minimal() +
  scale_x_date(date_breaks = "5 years",
               date_labels = "%Y") +
  labs(title = "Agulhas cross-shelf gradient", 
       subtitle = "Central Agulhas Bank Offshore SST - Central Agulhas Bank Coastal SST",
       x = "",
       y = "")
# ag_plot

pres_plot <- ggplot(pressure,aes(x = date, y = pressure), show.legend = FALSE) +
  geom_path()+
  geom_point()+
  theme_minimal() +
  scale_x_date(date_breaks = "5 years",
               date_labels = "%Y") +
  labs(title = "Atmospheric pressure (hPa)",
       subtitle = "NCEP/NCAR Reanalysis",
       x = "",
       y = "")
# pres_plot

cap_plot <- ggplot(cap_anomaly) +
  geom_bar(aes(x = date, y = cap_anomaly, fill = col), stat = "identity", show.legend = FALSE)+
  geom_hline(yintercept = 0, linewidth = .75) +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_minimal() +
  scale_x_date(date_breaks = "5 years",
               date_labels = "%Y") +
  scale_fill_manual(values = c("blue", "red")) +
  labs(title = "Standardized Cape Agulhas Pressure (CAP) anomaly",
       x = "",
       y = "")
# cap_plot

tsa_plot <- ggplot(tsa, aes(x = year, y = proportion)) +
  geom_path() +
  geom_point() +
  theme_minimal() +
    labs(x = "", y = "",
         title = "Thermal Spawning Area index",
         subtitle = "proportion of Agulhas Bank SST within 16–19°C during September and October") +
  NULL
# tsa_plot

sst_wide <- summer_sst %>% 
  pivot_wider(names_from = variable, values_from = mean)

sst_plot <- ggplot(summer_sst, aes(x = year, y = mean, color = variable)) +
  geom_path() +
  geom_point() +
  theme_minimal() +
    labs(x = "", y = "",
         title = "Mean temperature (°C)",
         subtitle = "December-February OISST") +
  NULL
# sst_plot

```
* Upwelling 
    + From Lamont et al 2018. Cumulative coastal South Benguela and Agulhas Bank upwelling from December to March
```{r, upwelling_dat}

# upwelling_month <- upwelling %>% 
#   group_by(area) %>%
#   mutate(upwelling_date = lag(year, n = 6), ##  years run from July (previous year) to June (current year). eg: 1979 = 1 July 1979 to 30 June 1980
#          upwelling_year = lubridate::year(upwelling_date),
#          months = lubridate::month(year)) %>% 
#   filter(!is.na(upwelling_date)) 
# 
# upwelling_dat <- upwelling_month %>% ## cumulative coastal upwelling from December to March
#   group_by(area, upwelling_year) %>%
#   summarize(value = sum(value, na.rm = TRUE))
# 
# upwelling_summer_dat <- upwelling_month %>% 
#   filter(!is.na(upwelling_date), 
#          months %in% c(12, 1, 2),
#          area != "total") %>% ## cumulative coastal upwelling from December to March
#   group_by(area, upwelling_year) %>%
#   summarize(value = mean(value, na.rm = TRUE))


up_summer <- ggplot(upwelling, aes(x = year, y = value/1000, color = variable)) +
  geom_path() +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "", y = "", color = "",
       title = "Upwelling index",
       subtitle = bquote("December-February cumulative upwelling"~(m^3~s^-1~100~m^-1)~x~1000)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# up_summer

upwelling_wide <- upwelling %>% 
  pivot_wider(names_from = variable, values_from = value)

```
* Primary Production
    + Southern Benguela - 
        * summer mean chlorophyll a (mg/m3)
        * summer mean fractional contribution of microphytoplankton, nanophytoplankton, and picophytoplankton
    + Agulhas Bank - 
        * summer mean chlorophyll a (mg/m3)
        * summer mean fractional contribution of microphytoplankton, nanophytoplankton, and picophytoplankton
        * fluosurf - *in situ* Autumn mean chlorophyll a (mg m-3) near the surface (5 m depth)

```{r pp_dat}
pp_wide <- ltl_dat %>% 
  select(year, chla_SB, chla_AB, pico_SB, pico_AB, nano_SB, nano_AB, micro_SB, micro_AB)

pp_dat <- ltl_dat %>% 
  select(year, chla_SB, chla_AB) %>% 
  pivot_longer(-year, names_to = "name", values_to = "value") %>% 
  separate_wider_delim(cols = "name", delim = "_", names = c("name", "area")) %>% 
  mutate(area = ifelse(area == "AB", "Agulhas Bank", "Southern Benguela"))


pp_plot <- ggplot(pp_dat, aes(x = year, y = value, color = area)) +
  geom_path() +
    scale_x_continuous(limits = c(1978, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "", y = "", color = "",
       title = bquote("Seasonal mean chlorophyll"~italic(a)),
       subtitle = bquote("December-February"~(mg~m^-3))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# pp_plot

pp_size <- ltl_dat %>% 
  select(year, pico_SB, pico_AB, nano_SB, nano_AB, micro_SB, micro_AB) %>% 
  pivot_longer(-year, names_to = "name", values_to = "value") %>% 
  separate_wider_delim(cols = "name", delim = "_", names = c("name", "area")) %>% 
  mutate(area = ifelse(area == "AB", "Agulhas Bank", "Southern Benguela"))

pp_size_plot <- ggplot(pp_size, aes(x = year, y = value, color = name)) +
  facet_wrap(~area, nrow = 2) +
  geom_path() +
    scale_x_continuous(limits = c(1978, 2020), breaks = seq(1965, 2020, 5)) +
    scale_y_continuous(limits = c(0, 1)) +
  labs(x = "", y = "", color = "",
       title = bquote("Seasonal mean fractional contribution to chlorophyll"~italic(a)),
       subtitle = bquote("December-February"~(mg~m^-3))) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# pp_size_plot
```
* Zooplankton
    + Southern Benguela - 
        * Zooplankton samples collected in St Helena Bay between 1951 and 1996, specifically during austral autumn (April-June), when peak recruitment of anchovy and sardine occurs, were analysed in terms of abundance, species and size composition of the copepod community following a standard, quantitative protocol (see Verheye et al., 1998).
    + Agulhas Bank - 
        * total_mgC_m2 - Mean integrated total copepod biomass (mg C m-2) in the upper 200 m of the water column
        * Cal_mgC_m2 - Mean integrated biomass (mg C m-2) of Calanus agulhensis (all stages) in the upper 200 m of the water column
        * SmCal_mgC_m2 - Mean integrated biomass (mg C m-2) of small calanoid copepods (mainly Paracalanidae and Clausocalanidae) in the upper 200 m of the water column
        * Oithona_mgC_m2 - Mean integrated biomass (mg C m-2) of Oithonidae in the upper 200 m of the water column
        * Oncaea_mgC_m2 - Mean integrated biomass (mg C m-2) of Oncaeidae in the upper 200 m of the water column
```{r zoop_data}
zoop_wide <- ltl_dat %>% 
   select(year, copepodabund_SB,accs_SB, mdlgzp_SB, smallzp_SB, copepod_AB, calanus_AB, smcopepod_AB)


zoop_dat <- ltl_dat %>% 
  select(year, copepodabund_SB,accs_SB, mdlgzp_SB, smallzp_SB, copepod_AB, calanus_AB, smcopepod_AB) %>% 
  pivot_longer(-year, names_to = "name", values_to = "value") %>% 
  separate_wider_delim(cols = "name", delim = "_", names = c("name", "area"))

zoop_sb <- zoop_dat %>% 
  filter(area == "SB") %>% 
  mutate(name = case_when(name == "mdlgzp" ~ "zooplankton_ML",
                          name == "smallzp" ~ "zooplankton_S",
                          name == "copepodabund" ~ "copepod_all",
                          TRUE ~ name),
         unit = factor(case_when(grepl("zoop", name) ~ "biomass",
                          grepl("cope", name) ~ "abundance",
                          grepl("accs", name) ~ "accs",
                          TRUE ~ NA_character_), levels = c("biomass", "abundance", "accs"), 
                       labels = c("biomass~anomaly", bquote("copepod abundance"~log(~n~m^-2~+~1)), 
                                                            "ACCS~index~(mm)")))

zoop_sb_plot <- ggplot(zoop_sb, aes(x = year, y = value, color = name)) +
  facet_wrap(~unit, scale = "free_y", ncol = 1, labeller = label_parsed) +
  geom_line() +
  geom_point() + 
    scale_x_continuous(limits = c(1950, 2020), breaks = seq(1950, 2020, 10)) +
  labs(x = "", y = "", color = "",
       title = "Zooplankton indicators",
       subtitle = "Southern Benguela") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# zoop_sb_plot

zoop_ab <- zoop_dat %>% 
  filter(area == "AB") %>% 
  mutate(area = ifelse(area == "AB", "Agulhas Bank", "Southern Benguela"))

zoop_ab_plot <- ggplot(zoop_ab, aes(x = year, y = value, color = name)) +
  geom_path() +
  geom_point() + 
    scale_x_continuous(limits = c(1950, 2020), breaks = seq(1950, 2020, 10)) +
  labs(x = "", y = "", color = "",
       title = bquote("Mean integrated zooplankton biomass"~(mg~C~m^-2)),
       subtitle = "Agulhas Bank") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# zoop_ab_plot
```
* Small pelagic population dynamics
    + Recruitment surveys occur between April and June
    + Biomass surveys occur between October and November
    + log(S/R) 
```{r recruits, warning=FALSE, message=FALSE}

rec1 <- ggplot(recruitment, aes(x = year, y = value, group = species, color = species)) +
  geom_path() +
  scale_x_continuous(limits = c(1979, 2021), breaks = seq(1965, 2020, 5)) +
    scale_color_brewer(palette = "Set2") +
  labs(x = "",
       y = "",
       title = "small pelagic recruitment",
       subtitle = "autumn/winter (May/June) acoustic survey (billions)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# rec1
```
```{r biomass, warning=FALSE, message=FALSE}

bio1 <- ggplot(biomass, aes(x = year, y = value, group = species, color = species)) +
  facet_wrap(~ factor(area, levels = c("east", "west", "total")), ncol = 1) +
  geom_path() +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "",
       y = "",
       title = "small pelagic spawner biomass",
       subtitle = "spring/summer (October/November) acoustic survey (million tonnes)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL

```


```{r logspr, warning=FALSE, message=FALSE}
logrps <- biomass %>% 
  filter(tolower(species) != "roundeye herring",
         area == "total") %>% 
  select(-area, -indicator, biomass = value) %>% 
  left_join(
    recruitment %>% 
      filter(tolower(species) != "roundeye herring") %>% 
      select(year, species, recruits = value),
    by = join_by(year, species)) %>% 
  group_by(year, species) %>%  
  mutate(recruits = recruits * 1e9,
         biomass = biomass * 1000,
         rps = recruits/biomass,
         log_rps = log(recruits / biomass))

logrps_plot <- ggplot(logrps, aes(x = year, y = log_rps, group = species, color = species)) +
  geom_path() +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  labs(x = "",
       y = "",
       title = "small pelagic survival",
       subtitle = "log(recruit/biomass)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL
# logrps_plot

rps_wide <- logrps %>% 
  select(-recruits, -rps) %>% 
  pivot_wider(names_from = species, values_from = c(biomass, log_rps))
```

* Fishing
    + Small pelagic catches occur throughout the year (see supplemental . 
    
```{r pel_catch, warning=FALSE, message=FALSE}

sp_plot <- ggplot(catches, aes(x = year, y = value/1000, group = species, fill = species, color = species)) +
  geom_path() +
  # geom_bar(stat = "identity") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "", y = "",
       title = "small pelagic catches",
       subtitle = "tonnes (thousands)") +
  NULL

```


```{r sardine_month, fig.width=12}

sardine_d <- catches_m %>% 
  filter(species == "sardine") %>% 
  select(-species) %>% 
  pivot_longer(Jan:Dec, names_to = "month", values_to = "value") %>% 
  mutate(month = factor(month.name[match(.$month, month.abb)], levels = month.name[c(6:12, 1:5)]),
         year = factor(year),
         table = case_when(table == "1a" ~ "directed fishery west of Cape Agulhas",
                           table == "1b" ~ "directed fishery east of Cape Agulhas",
                           table == "1c" ~ "bycatch with anchovy fishery",
                           TRUE ~ NA_character_)) 


sard1 <- ggplot(sardine_d, aes(x = month, y = year, height = value, group = year, fill = table)) +
  ggridges::geom_density_ridges(stat = "identity", scale = 3, alpha = 0.4, show.legend = FALSE) +
  facet_wrap(~table, ncol = 3) +
  labs(x = "", y = "",
       title = "distribution of monthly sardine commercial catches") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(axis.text.y = element_text(vjust = 0),
        axis.text.x = element_text(angle = 45, vjust = .5)) +
  NULL
```
```{r anchovy_month}

anchovy_d <- catches_m %>% 
  filter(species == "anchovy") %>% 
  select(-species, -table) %>% 
  pivot_longer(Jan:Dec, names_to = "month", values_to = "value") %>% 
  mutate(month = factor(month.name[match(.$month, month.abb)], levels = month.name[c(6:12, 1:5)]),
         year = factor(year)) 


anch1 <- ggplot(anchovy_d, aes(x = month, y = year, height = value, group = year)) +
  ggridges::geom_density_ridges(stat = "identity", scale = 3, fill = "purple", alpha = 0.4, show.legend = FALSE) +
  # facet_wrap(~table, ncol = 3) +
  labs(x = "", y = "",
       title = "distribution of monthly anchovy commercial catches") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(axis.text.y = element_text(vjust = 0),
        axis.text.x = element_text(angle = 45, vjust = .5)) +
  NULL

```
```{r calendar_table}
month_ord <- month.name[c(7:12, 1:6)]
paste("'", cal_tab$process, "'", collapse = ", ", sep =)
process_ord <- c('Upwelling (SB)', 'Upwelling (AB)', 'Sea-surface temperature (AB)', 
                 'Thermal spawning area (AB)', 'Cross-shelf Temperature gradient (AB)', 'Cape Agulhas Pressure (AB)', 
                 'Chl a concentration (AB)', 'Chl a concentration (SB)', 'Zooplankton biomass (AB)', 
                 'Zooplankton abundance (SB)', 'Zooplankton composition (SB)', 'Sardine recruitment survey (SB)',
                 'Anchovy recruitment survey (SB)', 'Sardine biomass survey (AB)', 'Anchovy biomass survey (AB)',
                 'Sardine bycatch', 'Anchovy catch', 'Sardine catch')

seasons <- data.frame(xstart = seq(0.5, 9.5, 3), xend = seq(3.5, 12.5, 3), 
                      season = factor(c("winter", "spring", "summer", "autumn"),
                                   levels = c("winter", "spring", "summer", "autumn")))

all_cal <- cal_tab %>% 
  mutate(year = ifelse(month_start > month_end, 2019, 2020),
         month_end = as.Date(sprintf("2020-%s-01", month_end)),
         month_start = as.Date(sprintf("%s-%s-01", year, month_start)),
         process = factor(process, levels = process_ord)) %>% 
  group_by(process) %>% 
  tidyr::complete(month_start = seq(month_start, month_end, by = "month")) %>% 
  tidyr::fill(lag, .direction = "down") %>% 
  select(process, month = month_start, lag) %>% 
  mutate(month = factor(lubridate::month(month, label = TRUE, abbr = FALSE),
                        levels = month_ord))

timeline_plot <- ggplot() +
  geom_rect(data = seasons, aes(xmin = xstart, xmax = xend, 
                                ymin = -Inf, ymax = Inf, fill = season), alpha = 0.4) +
  geom_point(data = all_cal, aes(x = month, y = process, color = as.factor(lag)), size = 5) +
  scale_color_manual(values = c("black", "red", "blue")) +
  theme_minimal() +
  labs(x = "", y = "", 
       color = "lag",
       title = "Timeline for important Southern Benguela processes",
       subtitle = "Lags indicate events in the current year (T = 0) that are affected in year T+lag") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = .5))
# timeline_plot
```

* These data are aggregated according to their temporal occurrence

```{r dsem_data, warning=FALSE, message=FALSE}

sem_dat_long <- data.frame(year = 1950:2023) %>% 
  left_join(upwelling_wide) %>% 
  left_join(cap_anomaly %>% select(-date, -col, cap = cap_anomaly)) %>% 
  left_join(tsa %>% select(-area, -spawner_area, tsa = proportion)) %>% 
  left_join(agulhas_gradient %>% select(agulhas_gradient = sst_gradient, year)) %>% 
  left_join(sst_wide) %>% 
  left_join(pp_wide) %>% 
  left_join(zoop_wide) %>% 
  left_join(rps_wide) %>% 
  left_join(catches %>% 
              select(year, species, value) %>% 
              pivot_wider(names_from = species, names_glue = "{tolower(gsub(' ', '_', species))}_catches", values_from = value)) %>% 
  pivot_longer(-year, names_to = "var", values_to = "val") %>% 
  group_by(var) %>% 
## Remove roundeye for now
  mutate(val = scale(val, scale = TRUE, center = TRUE)[,1]) %>% 
  # filter(!stringr::str_detect(var, "roundeye|_east|_west"))
    filter(!stringr::str_detect(var, "roundeye|_total")) 

sem_dat <- sem_dat_long %>% 
  pivot_wider(names_from = var, values_from = val) 

zplot <- ggplot(sem_dat_long, aes(x = year, y = val, group = var, color = var)) +
  geom_path(show.legend = FALSE) +
  gghighlight::gghighlight(unhighlighted_params = list(alpha = 0.4)) +
  facet_wrap(~var, ncol = 3) +
  labs(x = "", y = "z-score", color = "",
       title = "Indicators used in dsem analysis") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  NULL

# zplot

# timeline_plot
```

## Causal modeling ---
* Initial conceptual model is taken from ODEMM/IEA process (Skein et al 2022 and Bornman et al 2024)
* Dynamic Structural Equation Model (Thorson et al 2024)


# Results ---

* Simultaneous effects
    + Upwelling has a significant positive relationship with large phytoplankton and a negative relationship with small phytoplankton.
    + Large and small phytoplankton have significant positive relationships with all zooplankton size classes, but the largest effect size is between large phytoplankton and macrozooplankton and small phytoplankton and mesozooplankton. 
    + Benguela Nino has a negative relationship with sardine recruits. 
* Lag 1 (X in time T affects Y in time T+1)
    + Microplankton have a significant positive relationship with both anchovy and sardine recruits
    + Sardine recruits have a significant positive relationship with sardine total biomass, whereas anchovy does not have a strong relationship between recruits and biomass
        + Anchovy catches don’t have a strong relationship with sardine biomass (maybe think about area/regional effects)
* Lag 3 (X in time T affects Y in time T+3)
    + Sardine biomass has a significant positive relationship with sardine catches

```{r dag_full, include=FALSE}
dag_full <- phylopath::DAG(pp_large~upwelling,
                           pp_small~upwelling,
                           macrozooplankton~pp_large,
                           mesozooplankton~pp_large,
                           microzooplankton~pp_large,
                           macrozooplankton~pp_small,
                           mesozooplankton~pp_small,
                           microzooplankton~pp_small,
                           anchovy_recruits~macrozooplankton,
                           anchovy_recruits~mesozooplankton,
                           anchovy_recruits~microzooplankton,
                           anchovy_biomass~anchovy_recruits,
                           anchovy_catch~anchovy_biomass,
                           sardine_recruits~macrozooplankton,
                           sardine_recruits~mesozooplankton,
                           sardine_recruits~microzooplankton,
                           sardine_recruits~sbni,
                           sardine_biomass~anchovy_catch,
                           sardine_biomass~sardine_recruits,
                           sardine_catch~sardine_biomass)

df_plot <- plot(dag_full, edge_width = 1, type="width",
                text_size = 4, show.legend = FALSE, box_x = 12) +
  scale_x_continuous(expand = c(0.05, 0.05)) +  
  labs(title = "directed acyclic graph - full") +
  NULL

df_plot
```
* Still need to figure out how to characterize the lags in the DAG

```{r sem_full}
colnames(sem_dat)
sem_dat_full <- sem_dat %>% 
  select(year, 
         upwelling_AB, upwelling_SB, tsa, cap, agulhas_gradient,## environmental drivers
         chla_SB, chla_AB, ## primary production
         pico_SB, pico_AB, nano_SB, nano_AB, micro_SB, micro_AB, ## phytoplankton size
         copepodabund_SB, accs_SB, mdlgzp_SB, smallzp_SB, copepod_AB, calanus_AB, smcopepod_AB, # zooplankton community     
         biomass_Anchovy, biomass_Sardine, log_rps_Anchovy, log_rps_Sardine,## fish
         anchovy_catches, sardine_catches) %>% ## removals
  as.data.frame()



## guidance from ??make_dsem_ram()
### X -> Y, 1, XtoY indicates that X in time T affects Y in time T+1)
### X -> Y, 0, XtoY indicates X affects Y simultaneously (lag=0) 

# Specify model
sem_simple <- "
  # Link, lag, param_name (NA = off), fixed value (optional, 1 if param is off)
  
  ## AB environment mediates primary production
  upwelling_AB -> chla_AB, 0, upwelling_to_ppAB
  sst_AB -> chla_AB, 0, sst_to_ppAB
  # chla_SB -> nano_SB, 0, pp_to_nanoSB
  # chla_SB -> pico_SB, 0, pp_to_picoSB
  # chla_SB -> micro_SB, 0, pp_to_microSB

  ## SB environment mediates primary production
  # upwelling_SB -> chla_SB, 0, upwelling_to_ppSB
  # sst_SB -> chla_SB, 0, sst_to_ppSB


  ## AB Trophic links between PP and zooplankton
  chla_SB -> mdlgzp_SB, 0, chla_to_largeZP
  chla_SB -> copepodabund_SB, 0, chla_to_copeopodabundSP
  chla_SB -> smallzp_SB, 0, chla_to_smallZP

  ## SB Trophic links between PP and zooplankton
  chla_SB -> mdlgzp_SB, 0, chla_to_largeZP
  chla_SB -> copepodabund_SB, 0, chla_to_copeopodabundSP
  chla_SB -> smallzp_SB, 0, chla_to_smallZP

  ## Trophic
  mdlgzp_SB -> log_rps_Anchovy, 0, largeZP_to_rpsA
  smallzp_SB -> log_rps_Sardine, 0, smallZP_to_rpsS
  copepodabund_SB -> log_rps_Anchovy, 0, copepodabund_to_rpsA
  chla_SB -> log_rps_Sardine, 0, chla_to_rpsS


  ## Stock recruit
  biomass_Anchovy -> log_rps_Anchovy, 0, biomass_to_rpsA
  biomass_Sardine -> log_rps_Sardine, 0, biomassS_to_rpsS
  
  ## Spawning effects 
  tsa -> biomass_Anchovy, 0, tsa_to_biomassA
  tsa -> biomass_Sardine, 0, tsa_to_biomassS
  cap -> biomass_Anchovy, 0, cap_to_biomassA
  cap -> biomass_Sardine, 0, cap_to_biomassS
  agulhas_gradient -> biomass_Anchovy, 0, gradient_to_biomassA
  agulhas_gradient -> biomass_Sardine, 0, gradient_to_biomassS

  
  
  ## AR
  upwelling_SB -> upwelling_SB, 1, AR1, 0.001
  chla_SB -> chla_SB, 1,  AR2, 0.001
  mdlgzp_SB -> mdlgzp_SB, 1, AR3, 0.001
  smallzp_SB -> smallzp_SB, 1, AR4, 0.001
  log_rps_Anchovy -> log_rps_Anchovy, 1, AR5, 0.001
  log_rps_Sardine -> log_rps_Sardine, 1, AR6, 0.001
  biomass_Anchovy ->  biomass_Anchovy, 1, AR7, 0.001
  biomass_Sardine -> biomass_Sardine, 1, AR8, 0.001
  cap -> cap, 1, AR9, 0.001
  tsa -> tsa, 1, AR10, 0.001
  copepodabund_SB -> copepodabund_SB, 1, AR11, 0.001
  agulhas_gradient -> agulhas_gradient, 1, AR12, 0.001

  # anchovy_biomass_west -> anchovy_biomass_west, 1, AR9, 0.001
  # sardine_biomass_east -> sardine_biomass_east, 1, AR10, 0.001
  # anchovy_biomass_east -> anchovy_biomass_east, 1, AR11, 0.001
  # anchovy_catches -> anchovy_catches, 1, AR12, 0.001
  
  
  # upwelling_SB -> log_rps_Sardine, 0, upwelling_to_rpsS
  # upwelling_SB -> log_rps_Anchovy, 0, upwelling_to_rpsA
  
  ## Spawner conditions/transport (Agulhas)
  # cap -> log_rps_Sardine, 0, upwelling_to_rpsS
  # upwelling_SB -> log_rps_Anchovy, 0, upwelling_to_rpsA
  
  
  ## Sardine 
  # macrozooplankton -> sardine_recruits, 1, zoopMacro_to_recruitsS
  # mesozooplankton -> sardine_recruits, 1, zoopMeso_to_recruitsS
  # microzooplankton -> sardine_recruits, 1, zoopMicro_to_recruitsS
  
  # tsa -> biomass_Sardine, 0, tsa_to_biomassS
  # cap -> biomass_Sardine, 0, cap_to_biomassS
  
  # sardine_recruits -> sardine_biomass_east, 1, recruitsA_to_biomassES
  # anchovy_catches -> sardine_biomass_east, 1, catchesA_to_biomassES
  # sardine_catches -> sardine_biomass_east, 3, catchesS_to_biomassES
  # sardine_recruits -> sardine_biomass_west, 1, recruitsA_to_biomassWS
  # anchovy_catches -> sardine_biomass_west, 1, catchesA_to_biomassWS
  # sardine_catches -> sardine_biomass_west, 3, catchesS_to_biomassWS
  
  ## Anchovy
  # macrozooplankton -> anchovy_recruits, 1, zoopMacro_to_recruitsA
  # mesozooplankton -> anchovy_recruits, 1, zoopMeso_to_recruitsA
  # microzooplankton -> anchovy_recruits, 1, zoopMicro_to_recruitsA
  
  # tsa -> biomass_Anchovy, 0, tsa_to_biomassA
  # tsa -> anchovy_biomass_east, 1, tsa_to_biomassEA
  # cap -> biomass_Anchovy, 0, cap_to_biomassA
  # cap -> sardine_biomass_east, 1, cap_to_biomassEA
  # agulhas_gradient -> anchovy_biomass_east, 1, agulhas_to_biomassEA
  
  # anchovy_recruits -> anchovy_biomass_east, 1, recruitsA_to_biomassEA
  # anchovy_catches -> anchovy_biomass_east, 1, catchesA_to_biomassEA  
  # anchovy_recruits -> anchovy_biomass_west, 1, recruitsA_to_biomassWA
  # anchovy_catches -> anchovy_biomass_west, 1, catchesA_to_biomassWA  
  
"
sem_dat_simple <- sem_dat %>% 
  select(year, upwelling_SB, cap, tsa, agulhas_gradient, chla_SB, mdlgzp_SB, smallzp_SB, copepodabund_SB,
       log_rps_Anchovy, log_rps_Sardine, biomass_Anchovy, biomass_Sardine) %>% 
   filter(!if_all(-year, is.na)) %>% 
  filter(year >= 1975) %>% 
  as.data.frame()

rownames(sem_dat_simple) <- sem_dat_simple$year
sem_dat_simple <- sem_dat_simple[,-1]


# Fit dsem
fit_simple <- dsem( sem = sem_simple,
            tsdata = ts(sem_dat_simple),
            family =  rep('fixed', ncol(sem_dat_simple)),
            control = dsem_control(use_REML = TRUE, quiet=TRUE))
ParHat_simple = fit_full$obj$env$parList()

summary_simple <- summary(fit_simple) %>% 
  mutate(sig = cut(p_value, breaks = c(-Inf, 0.01, 0.05, 0.10, Inf), 
                   labels = c("***", "**", "*", "NS")))
# AIC(fit_simple)

# AIC(fit_full)
# 1551.142
# 1548
# 1545
# 1543

```


```{r dsem_full_dag, fig.height = 12, fig.width = 15}

lags_simple <- unique(as.numeric(summary(fit_simple)$lag))
for(i in lags_simple){
  
  p1_simple <- NULL
  p2_simple <- NULL

  p1_simple <- plot( (as_fitted_DAG(fit_simple, lag = i)), edge_width = 1, type="width",
                      text_size = 4, show.legend = FALSE, box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("lag %s effects", i)) +
    NULL

  p1_simple <- ggplot_build(p1_simple)
  p1_simple$data[[1]]$edge_width <- 1
  
  
  p2_simple <- plot( (as_fitted_DAG(fit_simple, what="p_value", lag = i)), #edge_width = 2, 
                      type = "width",
                      text_size = 4, show.legend = FALSE, colors=c('black', 'black'), 
                      box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("Two-sided p-value (lag %s)", i)) +
    NULL
  
  
  p2_simple <- ggplot_build(p2_simple)
  p2_simple$data[[1]]$edge_width <- 1
  
  print(wrap_ggplot_grob(ggplot_gtable(p1_simple))/wrap_ggplot_grob(ggplot_gtable(p2_simple)))
}

```

```{r dsem_full_ts, warning=FALSE, message=FALSE}

# Timeseries plot
raw_dat_simple <- data.frame(year = rownames(sem_dat_simple), 
                         sem_dat_simple) %>% 
  pivot_longer(!year, names_to = "var", values_to = "value")

pred_model_simple <- data.frame(year = rownames(sem_dat_simple), 
                         ParHat_simple$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "predicted")

pred_sd_simple <- data.frame(year = rownames(sem_dat_simple),
                      as.list(fit_simple$sdrep,what="Std.")$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "SD")

pred_all_simple <- raw_dat_simple %>% 
  left_join(pred_model_simple, by = join_by(year, var)) %>% 
  left_join(pred_sd_simple, by = join_by(year, var)) %>% 
  mutate(lower = predicted - ifelse(is.na(SD), 0, SD),
         upper = predicted + ifelse(is.na(SD), 0, SD),
         year = as.numeric(year)) 

ts_plot_simple <- ggplot(pred_all_simple, aes(x = year, y = predicted, ymin = lower, ymax = upper, group = var)) +
  geom_path() +
  geom_ribbon(alpha = 0.2, color = "blue", fill = "blue") +
  geom_point(aes(x = year, y = value)) +
  facet_wrap(~var) +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  theme_minimal() +
  labs(x = "", y = "",
       title = "Estimated values",
       subtitle = "measurements (black dots), predicted values (blue lines), and 95% predictive intervals (\u00B11.96 SE)") +
  NULL
ts_plot_simple
```



# Discussion ---


------

## Supplemental plots

### Data 
```{r supplement_plots, include=TRUE}

ag_plot
pres_plot
cap_plot
tsa_plot
bni_m / bni_s
up_summer
pp_plot
zoop_plot
rec1
bio1
sp_plot
sard1
anch1

```


### Anchovy model ---
```{r sem_anchovy,eval=FALSE, echo = FALSE, message=FALSE, warning=FALSE}
sem_dat_anch <- sem_dat %>% 
  select(year, upwelling, pp_large, pp_small, macrozooplankton,
         mesozooplankton, microzooplankton, 
         anchovy_recruits, anchovy_biomass_total, anchovy_catches) %>% 
  as.data.frame()

rownames(sem_dat_anch) <- sem_dat_anch$year
sem_dat_anch <- sem_dat_anch[,-1]
# Z = ts(sem_dat_anch)

# family = rep('fixed', ncol(sem_dat_anch))

## guidance from ??make_dsem_ram()
### X -> Y, 1, XtoY indicates that X in time T affects Y in time T+1)
### X -> Y, 0, XtoY indicates X affects Y simultaneously (lag=0) 
# Specify model
sem_anch <- "
  # Link, lag, param_name (NA = off), fixed value (optional, 1 if param is off)
  
  upwelling -> pp_large, 0, upwelling_to_ppL
  upwelling -> pp_small, 0, upwelling_to_ppS
  pp_large -> macrozooplankton, 0, ppL_to_zoopMacro
  pp_small -> macrozooplankton, 0, ppS_to_zoopMacro
  pp_large -> mesozooplankton, 0, ppL_to_zoopMeso
  pp_small -> mesozooplankton, 0, ppS_to_zoopMeso
  pp_large -> microzooplankton, 0, ppL_to_zoopMicro
  pp_small -> microzooplankton, 0, ppS_to_zoopMicro
  macrozooplankton -> anchovy_recruits, 1, zoopMacro_to_recruitsA
  mesozooplankton -> anchovy_recruits, 1, zoopMeso_to_recruitsA
  microzooplankton -> anchovy_recruits, 1, zoopMicro_to_recruitsA
  anchovy_recruits -> anchovy_biomass_total, 1, recruitsA_to_biomssA
  anchovy_biomass_total -> anchovy_catches, 1, biomassA_to_catches
  
  upwelling -> upwelling, 1, AR1, 0.001
  pp_large -> pp_large, 1,  AR2, 0.001
  pp_small -> pp_small, 1, AR3, 0.001
  macrozooplankton -> macrozooplankton, 1, AR4, 0.001
  mesozooplankton -> mesozooplankton, 1, AR5, 0.001
  microzooplankton -> microzooplankton, 1, AR6, 0.001
  anchovy_recruits ->  anchovy_recruits, 1, AR7, 0.001
  anchovy_biomass_total -> anchovy_biomass_total, 1, AR8, 0.001
  anchovy_catches -> anchovy_catches, 1, AR9, 0.001
"

# Fit dsem
fit_anch = dsem(sem = sem_anch,
            tsdata = ts(sem_dat_anch),
            family = rep('fixed', ncol(sem_dat_anch)),
            control = dsem_control(use_REML = TRUE, quiet=TRUE))

ParHat_anch = fit_anch$obj$env$parList()

```

```{r dag_anchovy, eval=FALSE}

dag_anchovy <- phylopath::DAG(pp_large~upwelling,
                    pp_small~upwelling,
                    macrozooplankton~pp_large,
                    mesozooplankton~pp_large,
                    microzooplankton~pp_large,
                    macrozooplankton~pp_small,
                    mesozooplankton~pp_small,
                    microzooplankton~pp_small,
                    anchovy_recruits~macrozooplankton,
                    anchovy_recruits~mesozooplankton,
                    anchovy_recruits~microzooplankton,
                    anchovy_biomass~anchovy_recruits,
                    anchovy_catch~anchovy_biomass)
  
da_plot <- plot(dag_anchovy, edge.width = 1, type="width",
           text_size = 5, show.legend = FALSE, box_x = 30) +
     scale_x_continuous(expand = c(0.4, 0.1))+
  labs(title = "directed acyclic graph - anchovy") +
  NULL
da_plot$layers[[1]]$mapping$edge_width = 1
da_plot

```

```{r dsem_anch_ts, eval=FALSE, warning=FALSE, message=FALSE}

# Timeseries plot
raw_dat <- data.frame(year = rownames(sem_dat_anch), 
                         sem_dat_anch) %>% 
  pivot_longer(!year, names_to = "var", values_to = "value")

pred_model <- data.frame(year = rownames(sem_dat_anch), 
                         ParHat_anch$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "predicted")

pred_sd <- data.frame(year = rownames(sem_dat_anch),
                      as.list(fit_anch$sdrep,what="Std.")$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "SD")

pred_all <- raw_dat %>% 
  left_join(pred_model, by = join_by(year, var)) %>% 
  left_join(pred_sd, by = join_by(year, var)) %>% 
  mutate(lower = predicted - ifelse(is.na(SD), 0, SD),
         upper = predicted + ifelse(is.na(SD), 0, SD),
         year = as.numeric(year)) 

ts_plot_anch <- ggplot(pred_all, aes(x = year, y = predicted, ymin = lower, ymax = upper, group = var)) +
  geom_path() +
  geom_ribbon(alpha = 0.2, color = "blue", fill = "blue") +
  geom_point(aes(x = year, y = value)) +
  facet_wrap(~var) +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  theme_minimal() +
  labs(x = "", y = "",
       title = "Estimated values",
       subtitle = "measurements (black dots), predicted values (blue lines), and 95% predictive intervals (\u00B11.96 SE)") +
  NULL
ts_plot_anch
```

```{r dsem_anch_dag, eval=FALSE, fig.height = 12, fig.width = 15}

anch_lags <- unique(as.numeric(summary(fit_anch)$lag))
for(i in anch_lags){
  
  p1_anch <- NULL
  p2_anch <- NULL

  p1_anch <- plot( (as_fitted_DAG(fit_anch, lag = i)), edge_width = 1, type="width",
                      text_size = 4, show.legend = FALSE, box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("lag %s effects", i)) +
    NULL

  p1_anch <- ggplot_build(p1_anch)
  p1_anch$data[[1]]$edge_width <- 1
  
  
  p2_anch <- plot( (as_fitted_DAG(fit_anch, what="p_value", lag = i)), #edge_width = 2, 
                      type = "width",
                      text_size = 4, show.legend = FALSE, colors=c('black', 'black'), 
                      box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("Two-sided p-value (lag %s)", i)) +
    NULL
  
  p2_anch <- ggplot_build(p2_anch)
  p2_anch$data[[1]]$edge_width <- 1
  
  print(wrap_ggplot_grob(ggplot_gtable(p1_anch))/wrap_ggplot_grob(ggplot_gtable(p2_anch)))
}

# 
# p1_anch <- plot( (as_fitted_DAG(fit_anch)), edge.width = 1, type="width",
#            text_size = 5, show.legend = FALSE, box_x = 30) +
#      scale_x_continuous(expand = c(0.4, 0.1))+
#   labs(title = "simultaneous effects")
# p1_anch$layers[[1]]$mapping$edge_width = 1
# p2_anch <- plot( (as_fitted_DAG(fit_anch, what="p_value")), edge.width = .5, type = "width",
#            text_size = 5, show.legend=FALSE, colors=c('black', 'black'), box_x = 30) +
#      scale_x_continuous(expand = c(0.4, 0.1)) +
#   labs(title = "Two-sided p-value")
# p2_anch$layers[[1]]$mapping$edge_width = 0.5
# 
# 

# 
# p1_anch
# p2_anch

```

```{r coef_anch, eval=FALSE}

coef_dat_anch <- data.frame(name = colnames(sem_dat_anch), summary(fit_anch$sdrep)[1:9,]) %>%
  rename(coef = 2, "std_error" = 3) %>%
  mutate(upper = coef + std_error,
         lower = coef - std_error)

coef_plot_anch <- ggplot(coef_dat_anch, aes(x = name, y = coef, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 0) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  labs(x = "", y = "standardized regression coefficient") +
  theme_bw()

coef_plot_anch
```

### Sardine model--- 

```{r sem_sardine, eval=FALSE, echo = FALSE}
sem_dat_sardine <- sem_dat %>% 
  select(year, 
         upwelling, sbni, ## environmental drivers
         pp_large, pp_small, ## primary producers
         macrozooplankton, mesozooplankton, microzooplankton, ## zooplankton
         sardine_recruits, sardine_biomass_total, ## fish
         anchovy_catches, sardine_catches) %>% ## removals
  as.data.frame()

rownames(sem_dat_sardine) <- sem_dat_sardine$year
sem_dat_sardine <- sem_dat_sardine[,-1]

## guidance from ??make_dsem_ram()
### X -> Y, 1, XtoY indicates that X in time T affects Y in time T+1)
### X -> Y, 0, XtoY indicates X affects Y simultaneously (lag=0) 

# Specify model
sem_sard <- "
  # Link, lag, param_name (NA = off), fixed value (optional, 1 if param is off)
  
  upwelling -> pp_large, 0, upwelling_to_ppL
  upwelling -> pp_small, 0, upwelling_to_ppS
  pp_large -> macrozooplankton, 0, ppL_to_zoopMacro
  pp_small -> macrozooplankton, 0, ppS_to_zoopMacro
  pp_large -> mesozooplankton, 0, ppL_to_zoopMeso
  pp_small -> mesozooplankton, 0, ppS_to_zoopMeso
  pp_large -> microzooplankton, 0, ppL_to_zoopMicro
  pp_small -> microzooplankton, 0, ppS_to_zoopMicro
  sbni -> sardine_recruits, 0, sbni_to_recruitsS
  macrozooplankton -> sardine_recruits, 1, zoopMacro_to_recruitsS
  mesozooplankton -> sardine_recruits, 1, zoopMeso_to_recruitsS
  microzooplankton -> sardine_recruits, 1, zoopMicro_to_recruitsS
  sardine_recruits -> sardine_biomass_total, 1, recruitsA_to_biomssS
  anchovy_catches -> sardine_biomass_total, 1, catchesA_to_biomassS
  sardine_biomass_total -> sardine_catches, 2, biomassA_to_catchesS
  
  upwelling -> upwelling, 1, AR1, 0.001
  pp_large -> pp_large, 1,  AR2, 0.001
  pp_small -> pp_small, 1, AR3, 0.001
  macrozooplankton -> macrozooplankton, 1, AR4, 0.001
  mesozooplankton -> mesozooplankton, 1, AR5, 0.001
  microzooplankton -> microzooplankton, 1, AR6, 0.001
  sardine_recruits ->  sardine_recruits, 1, AR7, 0.001
  sardine_biomass_total -> sardine_biomass_total, 1, AR8, 0.001
  anchovy_catches -> anchovy_catches, 1, AR9, 0.001
  sbni -> sbni, 1, AR10, 0.001
  sardine_catches -> sardine_catches, 1, AR11, 0.001
"

# Fit dsem
fit_sardine = dsem( sem = sem_sard,
            tsdata = ts(sem_dat_sardine),
            family =  rep('fixed', ncol(sem_dat_sardine)),
            control = dsem_control(use_REML = TRUE, quiet=TRUE))

ParHat_sardine = fit_sardine$obj$env$parList()


```

```{r dag_sardine, eval=FALSE}
dag_sardine <- phylopath::DAG(pp_large~upwelling,
                              pp_small~upwelling,
                              macrozooplankton~pp_large,
                              mesozooplankton~pp_large,
                              microzooplankton~pp_large,
                              macrozooplankton~pp_small,
                              mesozooplankton~pp_small,
                              microzooplankton~pp_small,
                              sardine_recruits~macrozooplankton,
                              sardine_recruits~mesozooplankton,
                              sardine_recruits~microzooplankton,
                              sardine_recruits~sbni,
                              sardine_biomass~anchovy_catches,
                              sardine_biomass~sardine_recruits,
                              sardine_catch~sardine_biomass)

ds_plot <- plot(dag_sardine,edge_width = 1, type="width",
                text_size = 4, show.legend = FALSE, box_x = 12) +
  scale_x_continuous(expand = c(0.05, 0.05)) +  
  labs(title = "directed acyclic graph - sardine") +
  NULL

ds_plot <- ggplot_build(ds_plot)
ds_plot$data[[1]]$edge_width <- 1
plot(ggplot_gtable(ds_plot))

```

```{r dsem_sardine_ts, eval=FALSE, warning=FALSE, message=FALSE}

# Timeseries plot
raw_dat_sardine <- data.frame(year = rownames(sem_dat_sardine), 
                         sem_dat_sardine) %>% 
  pivot_longer(!year, names_to = "var", values_to = "value")

pred_model_sardine <- data.frame(year = rownames(sem_dat_sardine), 
                         ParHat_sardine$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "predicted")

pred_sd_sardine <- data.frame(year = rownames(sem_dat_sardine),
                      as.list(fit_sardine$sdrep,what="Std.")$x_tj) %>% 
  pivot_longer(!year, names_to = "var", values_to = "SD")

pred_all_sardine <- raw_dat_sardine %>% 
  left_join(pred_model_sardine, by = join_by(year, var)) %>% 
  left_join(pred_sd_sardine, by = join_by(year, var)) %>% 
  mutate(lower = predicted - ifelse(is.na(SD), 0, SD),
         upper = predicted + ifelse(is.na(SD), 0, SD),
         year = as.numeric(year)) 

ts_plot_sardine <- ggplot(pred_all_sardine, aes(x = year, y = predicted, ymin = lower, ymax = upper, group = var)) +
  geom_path() +
  geom_ribbon(alpha = 0.2, color = "blue", fill = "blue") +
  geom_point(aes(x = year, y = value)) +
  facet_wrap(~var) +
  scale_x_continuous(limits = c(1979, 2020), breaks = seq(1965, 2020, 5)) +
  theme_minimal() +
  labs(x = "", y = "",
       title = "Estimated values",
       subtitle = "measurements (black dots), predicted values (blue lines), and 95% predictive intervals (\u00B11.96 SE)") +
  NULL
ts_plot_sardine
```

```{r dsem_sardine_dag, eval=FALSE, fig.height = 12, fig.width = 15}

sard_lags <- unique(as.numeric(summary(fit_sardine)$lag))
for(i in 0:sard_lags){
  
  p1_sardine <- NULL
  p2_sardine <- NULL

  p1_sardine <- plot( (as_fitted_DAG(fit_sardine, lag = i)), edge_width = 1, type="width",
                      text_size = 4, show.legend = FALSE, box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("lag %s effects", i)) +
    NULL

  p1_sardine <- ggplot_build(p1_sardine)
  p1_sardine$data[[1]]$edge_width <- 1
  
  
  p2_sardine <- plot( (as_fitted_DAG(fit_sardine, what="p_value", lag = i)), #edge_width = 2, 
                      type = "width",
                      text_size = 4, show.legend = FALSE, colors=c('black', 'black'), 
                      box_x = 12) +
    scale_x_continuous(expand = c(0.05, 0.05)) +
    labs(title = sprintf("Two-sided p-value (lag %s)", i)) +
    NULL
  
  
  p2_sardine <- ggplot_build(p2_sardine)
  p2_sardine$data[[1]]$edge_width <- 1
  
  print(wrap_ggplot_grob(ggplot_gtable(p1_sardine))/wrap_ggplot_grob(ggplot_gtable(p2_sardine)))
}

```

```{r sard_coef, eval=FALSE}

coef_dat_sardine <- data.frame(name = colnames(sem_dat_sardine), summary(fit_sardine$sdrep)[1:11,]) %>% 
  rename(coef = 2, "std_error" = 3) %>% 
  mutate(upper = coef + std_error,
         lower = coef - std_error)

coef_plot_sardine <- ggplot(coef_dat_sardine, aes(x = name, y = coef, ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 0) +
  geom_linerange() +
  geom_point() + 
  coord_flip() +
  labs(x = "", y = "standardized regression coefficient") +
  theme_bw()

# p1_sardine
# p2_sardine
coef_plot_sardine


```

```{r, eval = FALSE}
sem_full <- "
  # Link, lag, param_name (NA = off), fixed value (optional, 1 if param is off)
  
  ## Environment mediates primary production
  upwelling_SB -> chla_SB, 0, upwelling_to_ppSB

  ## Trophic links between PP and zooplankton
  # chla_SB -> copepodabund_SB, 0, chla_to_copepod
  chla_SB -> mdlgzp_SB, 0, chla_to_largeZP
  chla_SB -> smallzp_SB, 0, chla_to_smallZP

  ## Larval conditions/survival (Southern Benguela)
  # upwelling_SB -> log_rps_Sardine, 0, upwelling_to_rpsS
  # upwelling_SB -> log_rps_Anchovy, 0, upwelling_to_rpsA
  
  ## Spawner conditions/transport (Agulhas)
  # cap -> log_rps_Sardine, 0, upwelling_to_rpsS
  # upwelling_SB -> log_rps_Anchovy, 0, upwelling_to_rpsA
  
  
  ## Sardine 
  # macrozooplankton -> sardine_recruits, 1, zoopMacro_to_recruitsS
  # mesozooplankton -> sardine_recruits, 1, zoopMeso_to_recruitsS
  # microzooplankton -> sardine_recruits, 1, zoopMicro_to_recruitsS
  
  # tsa -> biomass_Sardine, 0, tsa_to_biomassS
  # cap -> biomass_Sardine, 0, cap_to_biomassS
  
  # sardine_recruits -> sardine_biomass_east, 1, recruitsA_to_biomassES
  # anchovy_catches -> sardine_biomass_east, 1, catchesA_to_biomassES
  # sardine_catches -> sardine_biomass_east, 3, catchesS_to_biomassES
  # sardine_recruits -> sardine_biomass_west, 1, recruitsA_to_biomassWS
  # anchovy_catches -> sardine_biomass_west, 1, catchesA_to_biomassWS
  # sardine_catches -> sardine_biomass_west, 3, catchesS_to_biomassWS
  
  ## Anchovy
  # macrozooplankton -> anchovy_recruits, 1, zoopMacro_to_recruitsA
  # mesozooplankton -> anchovy_recruits, 1, zoopMeso_to_recruitsA
  # microzooplankton -> anchovy_recruits, 1, zoopMicro_to_recruitsA
  
  # tsa -> biomass_Anchovy, 0, tsa_to_biomassA
  # tsa -> anchovy_biomass_east, 1, tsa_to_biomassEA
  # cap -> biomass_Anchovy, 0, cap_to_biomassA
  # cap -> sardine_biomass_east, 1, cap_to_biomassEA
  # agulhas_gradient -> anchovy_biomass_east, 1, agulhas_to_biomassEA
  
  # anchovy_recruits -> anchovy_biomass_east, 1, recruitsA_to_biomassEA
  # anchovy_catches -> anchovy_biomass_east, 1, catchesA_to_biomassEA  
  # anchovy_recruits -> anchovy_biomass_west, 1, recruitsA_to_biomassWA
  # anchovy_catches -> anchovy_biomass_west, 1, catchesA_to_biomassWA  
  
  ## AR
  upwelling -> upwelling, 1, AR1, 0.001
  pp_large -> pp_large, 1,  AR2, 0.001
  pp_small -> pp_small, 1, AR3, 0.001
  macrozooplankton -> macrozooplankton, 1, AR4, 0.001
  mesozooplankton -> mesozooplankton, 1, AR5, 0.001
  microzooplankton -> microzooplankton, 1, AR6, 0.001
  anchovy_recruits ->  anchovy_recruits, 1, AR7, 0.001
  sardine_biomass_west -> sardine_biomass_west, 1, AR8, 0.001
  anchovy_biomass_west -> anchovy_biomass_west, 1, AR9, 0.001
  sardine_biomass_east -> sardine_biomass_east, 1, AR10, 0.001
  anchovy_biomass_east -> anchovy_biomass_east, 1, AR11, 0.001
  anchovy_catches -> anchovy_catches, 1, AR12, 0.001
  # cap -> cap, 1, AR13, 0.001
  # tsa -> tsa, 1, AR14, 0.001
  # agulhas_gradient -> agulhas_gradient, 1, AR15, 0.001
"

```